Chaotic mixing of tidal debris

### 2015-06-08 ms.tex
added table of parameters for potential

\title{Chaotic diffusion of tidal streams}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
	    Kathryn V. Johnston\altaffilmark{\colum},
	    Monica Valluri\altaffilmark{\mich},
	    Sarah Pearson\altaffilmark{\colum},
	    Andreas H. W. K\"upper\altaffilmark{\colum},
	    David W. Hogg\altaffilmark{\nyu,\cds,\mpia}}
\date{\centering \today}

@@

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or ``Perfect Ellipsoid'' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential and are centrophilic.

@@

@@

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion. [... how and where to end this section?]
%
%Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.

@@

@@

In this work, we use the triaxial potential expression from \citet{leesuto03}, parametrized in a slightly different manner. In terms of spherical coordinates\footnote{$(r,\phi,\theta)$ = (radius, azimuth, colatitude)} with the radius normalized by the scale radius, $u = r/r_s$
\begin{align}
	\Phi(u,\phi,\theta) &\approx \frac{v_c^2}{A}\left[F_1(u) + \frac{1}{2}(e_b^2 + e_c^2)F_2(u) + \frac{1}{2} [(e_b\sin\theta \sin\phi)^2 + (e_c\cos\theta)^2] F_3(u) \right]\\
	A &= \left(\ln2 - \frac{1}{2}\right) + \left(\ln2-\frac{3}{4}\right) (e_b^2 + e_c^2)
\end{align}
where $e_b = \sqrt{1 - (b/a)^2}$, $e_c = \sqrt{1 - (c/a)^2}$, and $v_c$ is the circular velocity at the scale radius, $r_s$, for the spherical case. The functions $F_i(u)$ are given in the appendix of \cite{leesuto03}. We chose $r_s=20~{\rm kpc}$ and $v_c = 175~{\rm km}~{\rm s}^{-1}$ by taking the mean halo concentration for a ${\rm M}_{vir} \approx 10^{12}~\msun$ halo, $c_e\approx5$, from \cite{jing02} and by assuming $R_{vir}\approx200~{\rm kpc}$. Figure~\ref{fig:potential} shows equipotential contours of this potential in projection.

@@

As a demonstration of the small-scale frequency modulation, we consider the three orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:three-orbits}). To resolve the short-time behavior of the frequency diffusion (corresponding to motion across resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 35 orbital periods and shift the window by half an orbital period between each calculation of the most significant frequencies --- this is the smallest window size we could choose before NAFF begins to consistently fail to recover the frequencies for the strongly chaotic orbit.  Figure~\ref{fig:three-orbits-freqs} shows plots of the frequency ratios as computed in each window of time --- plotted are the percent deviations of the frequency ratios from the median value over all windows. The difference between the first window (green +) and the last window (red x) is 50 orbital periods. The frequencies of the weakly chaotic orbit (panel labeled $W$) evolve quickly but are bounded to a small volume with fractional size $\approx$0.3\% (presumably by nearby resonant surfaces). This is comparable to the frequency spread in globular cluster tidal debris (0.1--0.5\%).

@@

>>>>

\title{Chaotic mixing of tidal debris}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
	    Kathryn V. Johnston\altaffilmark{\colum},
	    Monica Valluri\altaffilmark{\mich},
	    Sarah Pearson\altaffilmark{\colum},
	    Andreas H. W. K\"upper\altaffilmark{\colum},
	    David W. Hogg\altaffilmark{\nyu,\cds,\mpia}}
\date{\centering \today}

@@

In order to explore the question of how chaos manifests over timescales much shorter than that predicted from generic chaos indicators, we must first understand the structure and behavior of orbits in gravitational potentials.

\subsection{Orbits in integrable potentials}

@@

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- for example, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}. In a globally integrable potential, the Hamiltonian may be written solely in terms of the actions, $H = H(\boldsymbol{J})$. Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or ``Perfect Ellipsoid'' model \citep[e.g.,][]{kuzmin73, deZeeuw85}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. The tube orbits preserve a sense of rotation about either the long or short axis and are therefore centrophobic, whereas the box orbits may approach arbitrarily close to the center of the potential (centrophilic) and may change the sense of rotation about any axis.

@@

\subsection{Orbits in non-integrable potentials}

[TODO: take cartoons from talk and include here for description]

@@

When $\epsilon$ is large, there is no general prediction for the resulting behavior, however it seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}. We therefore expect a large number of regular orbits will survive --- the so-called KAM tori --- however the tori that survive will be separated by regions of chaotic motion. Any transformations to angle-action coordinates must be defined locally due to the destruction of tori caused by the onset of chaotic motion. [... how and where to end this section?]

\subsection{The behavior of chaotic orbits in non-integrable potentials}

[??]

@@

Our goal is to map the orbit structure of arbitrary potentials, with an emphasis on identifying the chaotic orbits and and understanding the evolution of these orbits over short times. In this section, we describe the methods we will use to detect and quantify the strength of chaos for large grids of orbits.

@@

In this work, we use the triaxial potential expression from \citet{leesuto03}, parametrized in a slightly different manner. In terms of spherical coordinates\footnote{$(r,\phi,\theta)$ = (radius, azimuth, colatitude)} with the radius normalized by the scale radius, $u = r/r_s$
\begin{align}
	\Phi(u,\phi,\theta) &\approx \frac{v_c^2}{A}\left[F_1(u) + \frac{1}{2}(e_b^2 + e_c^2)F_2(u) + \frac{1}{2} [(e_b\sin\theta \sin\phi)^2 + (e_c\cos\theta)^2] F_3(u) \right]\label{eq:potential}\\
	A &= \left(\ln2 - \frac{1}{2}\right) + \left(\ln2-\frac{3}{4}\right) (e_b^2 + e_c^2)
\end{align}
where $e_b = \sqrt{1 - (b/a)^2}$, $e_c = \sqrt{1 - (c/a)^2}$, and $v_c$ is the circular velocity at the scale radius, $r_s$, for the spherical case. The functions $F_i(u)$ are given in the appendix of \cite{leesuto03}. We chose $r_s=20~{\rm kpc}$ and $v_c = 175~{\rm km}~{\rm s}^{-1}$ by taking the mean halo concentration for a ${\rm M}_{vir} \approx 10^{12}~\msun$ halo, $c_e\approx5$, from \cite{jing02} and by assuming $R_{vir}\approx200~{\rm kpc}$. Figure~\ref{fig:potential} shows equipotential contours of this potential in projection, and Table~\ref{tbl:potential} summarizes the potential parameters.

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{ c  c }
	         Parameter & Value\\\toprule
		$v_c$ & 175~km~s$^{-1}$\\
		$r_s$ & 20~kpc\\
		$a$ & 1\\
		$b$ & 0.77\\
		$c$ & 0.55\\
		\bottomrule
		\end{tabular}
	\caption{Summary of potential parameters for the triaxial NFW potential (Equation~\ref{eq:potential}) used in this work. Triaxiality is introduced in the density to ensure the density is physical at all radii. Velocity scale, scale radius, and axis ratios are chosen to match the median halo parameters for a $M_{\rm vir} \approx 10^{12}~\msun$ halo from \citep{jing02}. \label{tbl:potential}}
\end{center}
\end{table*}

@@

As a demonstration of the small-scale frequency modulation, we consider again the three orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:three-orbits}). Figure~\ref{fig:fdiff-zooms} shows zoomed-reruns of the frequency diffusion time map of Figure~\ref{fig:freqdiff} centered on the three orbits --- the pixel positions are now shown relative to the three parent orbits. The frequency diffusion times shown in these grids were computed with two consecutive windows of length 35 orbital periods (shorter than the windows used in the previous, large grid of Figure~\ref{fig:freqdiff}). The resonant structure around the mildly chaotic orbit (white regions in panel W) bound the orbit to a small region, whereas for the strongly chaotic orbit (S), the entire region surrounding the orbit is strongly chaotic. To resolve the short-time behavior of the frequency diffusion (corresponding to motion across resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 35 orbital periods and shift the window by half an orbital period between each calculation of the most significant frequencies --- this is the smallest window size we could choose before NAFF begins to consistently fail to recover the frequencies for the strongly chaotic orbit.  Figure~\ref{fig:three-orbits-freqs} shows plots of the frequency ratios as computed in each window of time --- plotted are the percent deviations of the frequency ratios from the median value over all windows. The difference between the first window (green +) and the last window (red x) is 50 orbital periods. The frequencies of the weakly chaotic orbit (panel labeled $W$) evolve quickly but are bounded to a small volume with fractional size $\approx$0.3\% (presumably by nearby resonant surfaces). This is comparable to the frequency spread in globular cluster tidal debris (0.1--0.5\%).

@@

% Figure ??
\clearpage
\begin{figure*}[p]
\begin{center}
\includegraphics[width=1.1\textwidth]{figures/fdiff_zooms.png}
\caption{ } \label{fig:fdiff-zooms}
\end{center}
\end{figure*}
